---
title: '**Exploring the Legacy of Redlining on Environmental Justice in Los Angeles**'
author: "Yifei Liu"
date: '2023-10-17'
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Background

Present-day environmental justice may reflect legacies of injustice in the past. The United States has a long history of racial segregation which is still visible. During the 1930's the Home Owners' Loan Corporation (HOLC), as part of the New Deal, rated neighborhoods based on their perceived safety for real estate investment. Their ranking system, (A (green), B (blue), C (yellow), D (red)) was then used to block access to loans for home ownership. Colloquially known as "redlining", this practice has had widely-documented consequences not only for community wealth, but also health.^[Gee, G. C. (2008). A multilevel analysis of the relationship between institutional and individual racial discrimination and health status. American journal of public health, 98(Supplement_1), S48-S56.] Redlined neighborhoods have less greenery^[Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A. (2021). Redlines and greenspace: the relationship between historical redlining and 2010 greenspace across the United States. Environmental health perspectives, 129(1), 017006.] and are hotter than other neighborhoods.^[Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects of historical housing policies on resident exposure to intra-urban heat: a study of 108 US urban areas. Climate, 8(1), 12.] 

A recent study found that redlining has not only affected the environments communities are exposed to, it has also shaped our observations of biodiversity.^[Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical redlining is associated with increasing geographical disparities in bird biodiversity sampling in the United States. Nature Human Behaviour, 1-9.] Community or citizen science, whereby individuals share observations of species, is generating an enormous volume of data. Ellis-Soto and co-authors found that redlined neighborhoods remain the most undersampled areas across 195 US cities. This gap is highly concerning, because conservation decisions are made based on these data.

For further reading on the subject, refer to the coverage by the [New York Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html) and [EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

## Project Description

This project explores the impacts of historical redlining on current environmental justice issues in Los Angeles. It investigates how redlining — a discriminatory practice from the 1930s that graded neighborhoods for loan eligibility based on racial composition and other factors — has shaped current environmental disparities. This analysis brings together historical HOLC grade designations, environmental data from EJScreen, and biodiversity observations from the Global Biodiversity Information Facility.

## Project Goals

The primary objectives of this project are to:

- Analyze the relationship between historical redlining grades and current environmental indicators.
- Investigate biodiversity disparities in areas affected by redlining.
- Utilize spatial analysis and data visualization to present findings effectively.

## Skills Demonstrated

- **Spatial Data Handling**: Proficiency in managing and analyzing spatial data with `sf` and `tmap` in R.
- **Data Wrangling and Analysis**: Utilizing `tidyverse` tools for data cleaning and analysis.
- **Statistical Analysis**: Ability to derive meaning insights from statistical summaries and comparisons. 
- **Effective Visualization**: Creating informative and interpretive visualizations with `tmap` and `ggplot2`. 

## Data Sources and Preparation
### EJScreen

We will be working with data from the United States Environmental Protection Agency's EJScreen: Environmental Justice Screening and Mapping Tool. According to the [US EPA website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen), EJScreen provides on environmental and demographic information for the US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and [block group](https://en.wikipedia.org/wiki/Census_block_group) levels. This data helps assess current environmental disparities in Los Angeles County.  

We will be working with block group data that has been downloaded from the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data). To understand the associated data columns, we will need to explore the Technical Documentation and column description spreadsheet available in the data folder and the [limitations and caveats](https://www.epa.gov/ejscreen/limitations-and-caveats-using-ejscreen) of the data.

### Mapping Inequality

A team of researchers, led by the [Digital Scholarship Lab](https://dsl.richmond.edu/) at the University of Richmond have digitized maps and information from the HOLC as part of the [Mapping Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58) project. 

We will be working with maps of HOLC grade designations for Los Angeles. Information on the data can be found [here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).^[Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan Connolly, et al., “Mapping Inequality,” American Panorama, ed. Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023, https://dsl.richmond.edu/panorama/redlining/] These will serve as a basis for understanding historical redlining practices in Los Angeles.

### Biodiversity observations
Data from the [Global Biodiversity Information Facility](gbif.org)  on bird observations from 2021 onwards offers insights into current biodiversity patterns in Los Angeles. Global Biodiversity Information Facility is the largest aggregator of biodiversity observations in the world. Observations typically include a location and date that a species was observed. 


## Analysis Workflow

### Data import and Prepararion

#### Load relevant packages
In this section, we load necessary R libraries for spatial data manipulation, visualization, and general data processing. 
```{r package, message=FALSE, warning=FALSE}
# Load libraries
library(sf)
library(tmap)
library(tidyverse)
library(knitr)
library(ggplot2)
```

#### Datasets Reading 
We focus on three main datasets for this analysis: EJScreen environmental data, historical redlining data, and bird observation data. Each dataset is read and prepared for subsequent analysis.

```{r read data, message=FALSE, warning=FALSE}
# Read in EJScreen data and filter to LA
LA <- st_read("data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb/",
              query = "SELECT * FROM EJSCREEN_StatePctiles_with_AS_CNMI_GU_VI\
              WHERE CNTY_NAME = 'Los Angeles County'", 
              quiet = TRUE)

# Read in redlining data
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/citiesData/CALosAngeles1939/geojson.json",
              quiet = TRUE) %>%
  st_make_valid()

# Read in bird data
bird <- st_read("data/gbif-birds-LA", quiet = TRUE) %>% 
  filter(year == 2022)
```

### Investigate the legacy of redlining in current environmental (in)justice
#### Waste Water Discharge Analysis

In this section, we delve into the spatial distribution of wastewater discharge in Los Angeles County. The focus is on identifying census block groups with high wastewater discharge, particularly those exceeding the 95th percentile nationally. We will make a map of wastewater discharge by census block groups and indicate high wastewater discharge census block groups by adding a centroid.

```{r waste water, message=FALSE}
# Select wastewater data
wastewater <- LA %>% 
  select(ID, P_PWDIS)

# Filtering census block groups > 95th wastewater discharge
ab_wastewater <- wastewater %>% 
  filter(P_PWDIS > 95)

# Creating centroids for visualization
abww_centroid <- st_centroid(ab_wastewater)

# Mapping wastewater discharge
breaks_fill = c(0, 25, 50, 75, 95, 100)
tm_shape(wastewater) +
  tm_fill("P_PWDIS", 
          title = "Wastewater discharge Percentile",
          breaks = breaks_fill, 
          colorNA = "White",
          border.col = "#8080804D",
          border.lwd = 0.1) +
  tm_shape(abww_centroid) +
  tm_symbols(col = "P_PWDIS", 
             palette = "heat", 
             size = 0.5,
             title.col = "95 - 100 percentile Centroid",
             legend.show = TRUE) +
  tm_compass(type = "arrow", position = c("right","bottom"), size = 2) +
  tm_scalebar(position = c("right","bottom")) +
  tm_graticules(lines = FALSE)
```

#### Socioeconomic and Environmental Impact Analysis

In this section, we aim to examine the socioeconomic and environmental factors across census block groups in Los Angeles County. 

First we will evaluate the percentage of census block groups where less than 5% of the population is considered low income. This data helps in identifying areas with relatively higher economic status.

```{r cb1}
# Filtering data for low income population
cb1 <- LA %>% 
  filter(LOWINCPCT < 0.05) %>% 
  nrow()/nrow(LA)*100

# Displaying the percentage
print(paste("Percentage of census block groups:", cb1, "%"))
```

Next, we assess the proportion of census block groups that simultaneously rank high (above the 80th percentile) in particulate matter (PM 2.5) concentration and proximity to Superfund sites. This analysis helps in identifying areas with compounded environmental challenges.
```{r cb2}
# Filtering data for high environmental burden
cb2 <- LA %>% 
  filter(P_D2_PM25 > 80) %>% 
  filter(P_D2_PNPL > 80) %>% 
  nrow()/nrow(LA)*100

# Displaying the results
print(paste("Percentage of census block groups:", cb2, "%"))
```

#### Redlining Map Visualization

To visually comprehend the extent of historical redlining in Los Angeles, we'll create a map delineating historical redlining boundaries based on their HOLC grades. This visualization aids in understanding the geographical distribution of redlining's impact.

```{r HOLC map}
# Mapping redlining colored by HOLC grades
tm_shape(LA_redlining) +
  tm_fill("grade", title = "HOLC grade") +
  tm_borders() +
  tm_compass(type = "arrow", position = c("right","bottom"), size = 2) +
  tm_scalebar(position = c("right","bottom")) +
  tm_graticules(lines = FALSE)
```


#### Analyzing Census Block Groups within HOLC Grades

We will analyze census block groups overlap with historical redlining areas to understand how current census block groups correlate with historically redlined areas. 

```{r num_cb_in_rd}
# Transforming LA data to same CRS
LA <- st_transform(LA, crs = st_crs(LA_redlining))

# Filtering CBG in redlining
LA_in_rd <- LA %>% 
  st_filter(y = LA_redlining, .predicate = st_intersects)

# Displaying the number of affected census block groups
print(paste("Number of census block groups that fall within areas with HOLC grades:", 
          nrow(LA_in_rd)))

# Checking LA_in_rd by visualizing
tm_shape(LA) +
  tm_borders(col = "#8080804D") +
  tm_shape(LA_redlining) +
  tm_polygons(fill = "blue") +
  tm_shape(LA_in_rd) +
  tm_dots(fill = "red") +
  tm_compass(type = "arrow", position = c("right","bottom"), size = 2) +
  tm_scalebar(position = c("right","bottom")) +
  tm_graticules(lines = FALSE)
```

#### Summarizing Current Conditions within Redlining Categories

We aim to compare and contrast current environmental and socio-economic conditions across different redlining categories. We will summarize current conditions based on EJScreen data within historical redlining categories using the mean of the following variables:  
-% low income.  
- percentile for particulate Matter 2.5.  
- percentile for low life expectancy.  
- percentile for air toxics cancer risk

This summary will highlight disparities in environmental quality and socio-economic status influenced by historical redlining.

```{r summary}
# Joining CBG with redlining
LA_in_rd_join <- st_join(LA_in_rd, LA_redlining, join = st_intersects, left = FALSE)

# Generating summary statistics
summary <- st_drop_geometry(LA_in_rd_join) %>% 
  filter(!is.na(grade)) %>% 
  group_by(grade) %>% 
  summarise(
    low_income = mean(LOWINCPCT, na.rm = TRUE),
    p_pm_25 = mean(P_PM25, na.rm = TRUE),
    p_low_life_exp = mean(P_LIFEEXPPCT, na.rm = TRUE),
    p_air_cancer = mean(P_CANCER, na.rm = TRUE)
  )

# Displaying the summary table
kable(summary)
```

**The results from the summary table clearly reveals the environmental and socioeconomic disparities among the historical redlining categories. Census block groups with the highest redlining grade, "A", exhibit the best conditions with the lowest percentages of low income (14.9%), lowest PM 2.5 percentiles (72.1%), lowest low life expectancy percentile (23.7%), and lowest air toxics cancer risk percentile (44.0%). On the contrary, census block groups with the lowest redlining grade, "D", bears the worst conditions, demonstrating the highest percentages or percentiles across the variables (39.1%, 80.2%, 53.1%, 56.5%, respectively). This gradient is consistently observed across all categories, with areas labeled "B" and "C" facing intermediate challenges.**


### Investigate the legacy of redlining in biodiversity observations

#### Biodiversity Observations Analysis

This analysis focuses on understanding how redlining may have influenced current patterns of bird biodiversity in Los Angeles. By examining bird observations from 2022 within neighborhoods categorized by HOLC grades, we aim to unveil potential disparities in biodiversity stemming from historical urban planning decisions.

```{r summary bird}
# Transforming bird data to same CRS as LA_redlining
bird <- st_transform(bird, crs = st_crs(LA_redlining))

# Spatially joining bird observations with redlining data
bird_2022_in_rd <- st_join(bird, LA_redlining, join = st_intersects, left = FALSE)

# Calculating the total number of bird observations in redlined areas
num_bird_in_rd <- nrow(bird_2022_in_rd %>% filter(!is.na(grade)))

# Creating a summary table for bird observations by HOLC grades
summary_bird <- st_drop_geometry(bird_2022_in_rd) %>% 
  filter(!is.na(grade)) %>% 
  group_by(grade) %>% 
  summarise(
    count = n(),
    percent = count/num_bird_in_rd
  )

# Displaying the summary table
kable(summary_bird)
```

#### Visualizing of Bird Observation Data

To better visualize the distribution of bird observations across different redlining categories, we will create a histogram. This visual representation will highlight any significant disparities in bird biodiversity related to historical redlining grades.

```{r plot bird}
# Creating a histogram for bird observation percentages by HOLC grades
bird_plot <-ggplot(summary_bird,
                   aes(x = grade, y = percent, fill = grade)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(x = "HOLC Grade",
       y = "Bird Observation Percent",
       title = "Percent of 2022 Bird Obserations by HOLC Grades",
       fill = "HOLC Grade")

# Displaying the histogram
bird_plot
```

**The summary table and the bar plot illustrate the distribution of bird observations across the historical redlining categories in Los Angeles in 2022. Notably, "C" grade regions, not the expected "A" or "D" grades, lead with 35.7% of observations, while "A" grade regions have the fewest at 15.7%. It is surprising to see this pattern, as I expected more green spaces in higher-graded region, based on their superior environmental and socioeconomic scores in the previous summary table. This pattern suggest that bird abundance in LA is not strictly influenced by environmental and socioeconomic indicators computed in HOLC grades. Factors such as green spaces, habitat availability, and urbanization levels may contribute to these patterns, raising issue about the complex relationship between urbanization and biodiversity.**
